Countering language attrition with PanLex and the Web of Data

### 2014-03-08 panlex-rdf-paper/2012/MLOD_PanLex/paper_v5.tex
Except for the reference section which is too close to the footnotes, I guess my part is done.

\section{Linking}
\label{sec:linking}
The SML view in (\autoref{fig:ex:sparqlify-ml}) already establishes the interlinking of the PanLex languages with Lexvo.
In this section, we outline the interlinking with DBpedia, where we were interested in creating \emph{valid} and \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38}}, which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and removal of punctuation characters.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
\autoref{fig:plx-dbp-link-counts} summarizes the number of links obtained.

@@

\section{Publishing}
\label{sec:publishing}
With our RDF conversion work, we complement existing
APIs\footnote{\url{http://panlex.org/try/}} with Linked Data, powered by
Pubby\footnote{\url{http://wifo5-03.informatik.uni-mannheim.de/pubby/}},
and two SPARQL
endpoints\footnote{\url{http://ld.panlex.org/vsparql}}
\footnote{\url{http://ld.panlex.org/sparql}}, ran by Sparqlify and Virtuoso.
% and \url{http://ld.panlex.org/snorql}}.
An overview is shown in ~\autoref{fig:panlex-architecture}.
The SPARQL browser
\emph{SNORQL}\footnote{\url{https://github.com/kurtjx/SNORQL}} can be accessed
by replacing \emph{sparql} with \emph{snorql} in the respective links. Our SML
views and the interlinking code are hosted on GitHub\footnote{\url{https://github.com/AKSW/PanLex-2-RDF}}.
The created linksets are hosted in the PanLex database and are published together with the other data using Sparqlify.
Finally, we offer downloads tagged with timestamps of their creation\footnote{\url{http://ld.panlex.org/downloads/releases/}}.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{images/pdf/sparqlify_setup02.pdf}
\caption{PanLex architecture}
\label{fig:panlex-architecture}
\end{figure}

@@

\section{Related Work}
\label{sec:related}
PanLex is a project whose editors integrate information discovered from many lexical resources.
The extraction of information from linguistic sources, and techniques for automatically inferring translations,
are relevant work discussed in \cite{panlex_probtrans}.
%A similar approach of linking expressions to meanings is followed by the
% \emph{Global Wordnet Association}.
An important related initiative is the Global Wordnet Association (GWA)\footnote{\url{http://www.globalwordnet.org/}\vspace{0.7em}},
which offers a platform for sharing wordnets and, among other goals, uniformly representing wordnets
of different languages and establishing a universal index of meaning.
Wordnets are usually focused on the
definition of \emph{synsets} and relations between them in a single language; GWA is helping to transform these single-language synonym silos into a virtual multilingual translation resource. PanLex is approximating that, in a different way, by integrating data from numerous wordnets along with translingual sources into a single graph.
%As already mentioned in PanLex there can be several meaning entities
% representing the same unique sense.
In the Semantic Web context, several standard or quasi-standard vocabularies and ontologies have been developed with the rise of the \emph{Linguistic Linked Open Data} movement.
Examples include the
\emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010} for modeling
lexicon and machine-readable dictionaries, \emph{POWLA} for modeling linguistic corpora\cite{powla2012} and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

@@

We intend to address some limitations in the future:
The relations among PanLex's information sources, if treated as distinct datasets, could be modeled with the VoID vocabulary\footnote{\url{http://rdfs.org/ns/void}}.
The source entity should be refactored to reference users and information sources as distinct entities.
%Attribution is currently limited to (assertion) sources,
% obscuring the possible roles of multiple editors in iteratively interpreting sources and the subsidiary sources of data in sources that are %themselves aggregations.
Metadata attached to PanLex denotations are currently limited to arbitrary pairs of strings,
but this sacrifices discovery possibilities when the metadata
%(as they often do)
describe facts that can again be expressed with PanLex expressions.
% (e.g., that the expression is archaic).
Finally, new collaborations between PanLex and related fields (e.g. as language identification, language geolocation, lemmatization, transliteration, localization, etc.) are promising arenas for development.

>>>>

\section{Linking}
\label{sec:linking}
The SML view in (\autoref{fig:ex:sparqlify-ml}) establishes the interlinking of languages in PanLex with Lexvo.
Here we outline the interlinking with DBpedia, where we were interested in creating \emph{valid} and \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38}}, which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and removal of punctuation characters.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
\autoref{fig:plx-dbp-link-counts} summarizes the number of links obtained.

@@

\section{Publishing}
\label{sec:publishing}
With our RDF conversion work, we complement existing
APIs\footnote{\url{http://panlex.org/try/}} with Linked Data, powered by
Pubby\footnote{\url{http://wifo5-03.informatik.uni-mannheim.de/pubby/}},
and two SPARQL
endpoints\footnote{\url{http://ld.panlex.org/vsparql}}\textsuperscript{,}\footnote{\url{http://ld.panlex.org/sparql}}, ran by Sparqlify and Virtuoso.
% and \url{http://ld.panlex.org/snorql}}.
An overview is shown in ~\autoref{fig:panlex-architecture}.
The SPARQL browser
\emph{SNORQL}\footnote{\url{https://github.com/kurtjx/SNORQL}} can be accessed
by replacing \emph{sparql} with \emph{snorql} in the respective links. Our SML
views and the interlinking code are hosted on GitHub\footnote{\url{https://github.com/AKSW/PanLex-2-RDF}}.
The created linksets are hosted in the PanLex database and are published together with the other data using Sparqlify.
Finally, we offer downloads tagged with timestamps of their creation\footnote{\url{http://ld.panlex.org/downloads/releases/}}.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{images/pdf/sparqlify_setup02.pdf}
\caption{PanLex architecture}
\label{fig:panlex-architecture}
\end{figure}

@@

\section{Related Work}
\label{sec:related}
PanLex is a project whose editors integrate information discovered from many lexical resources.
The extraction of information from linguistic sources, and techniques for automatically inferring translations,
are relevant work discussed in \cite{panlex_probtrans}.
%A similar approach of linking expressions to meanings is followed by the
% \emph{Global Wordnet Association}.
An important related initiative is the Global Wordnet Association (GWA)\footnote{\url{http://www.globalwordnet.org/}},
which offers a platform for sharing wordnets and, among other goals, uniformly representing wordnets
of different languages and establishing a universal index of meaning.
Wordnets are usually focused on the
definition of \emph{synsets} and relations between them in a single language; GWA is helping to transform these single-language synonym silos into a virtual multilingual translation resource. PanLex is approximating that, in a different way, by integrating data from numerous wordnets along with translingual sources into a single graph.
%As already mentioned in PanLex there can be several meaning entities
% representing the same unique sense.
In the Semantic Web context, several standard or quasi-standard vocabularies and ontologies have been developed with the rise of the Linguistic LOD movement.
Examples include the
\emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010} for modeling
lexicon and machine-readable dictionaries, \emph{POWLA} for modeling linguistic corpora\cite{powla2012} and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

@@

\vfill
%\columnbreak

We intend to address some limitations in the future:
The relations among PanLex's information sources, if treated as distinct datasets, could be modeled with the VoID vocabulary\footnote{\url{http://rdfs.org/ns/void}}.
The source entity should be refactored to reference users and information sources as distinct entities.
%Attribution is currently limited to (assertion) sources,
% obscuring the possible roles of multiple editors in iteratively interpreting sources and the subsidiary sources of data in sources that are %themselves aggregations.
Metadata attached to PanLex denotations are currently limited to arbitrary pairs of strings,
but this sacrifices discovery possibilities when the metadata
%(as they often do)
describe facts that can again be expressed with PanLex expressions.
% (e.g., that the expression is archaic).
Finally, new collaborations between PanLex and related fields (e.g. as language identification, language geolocation, lemmatization, transliteration, localization, etc.) are promising areas for development.

\vspace{-10pt}
